\documentclass[12pt,fleqn]{exam}
\usepackage{pifont}
\usepackage{dingbat}
\usepackage{amsmath,amssymb} 
\usepackage{epsfig}
\usepackage[final]{hyperref}
\usepackage[letterpaper, margin=0.75in]{geometry}
\usepackage{tikz}
\usepackage[super]{nth}
\usetikzlibrary{arrows}
\addpoints
\boxedpoints
\pointsinmargin
\pointname{pts}

\usepackage{gensymb}
\usepackage{pdfpages}
\usepackage[final]{microtype}
\usepackage[american]{babel}
\usepackage[T1]{fontenc}
\usepackage{fourier}
%\usepackage{eucal}

\usepackage{isomath}
\usepackage{upgreek,amsmath}
\usepackage{amssymb}
\usepackage{siunitx}

\newcommand{\dotprod}{\, {\scriptzcriptztyle
    \stackrel{\bullet}{{}}}\,}

\newcommand{\reals}{\mathbf{R}}
\newcommand{\lub}{\mathrm{lub}} 
\newcommand{\glb}{\mathrm{glb}} 
\newcommand{\complex}{\mathbf{C}}
\newcommand{\dom}{\mbox{dom}}
\newcommand{\range}{\mbox{range}}
\newcommand{\cover}{{\mathcal C}}
\newcommand{\integers}{\mathbf{Z}}
%\newcommand{\degree}{\mathrm{degree}}
\newcommand{\vi}{\, \mathbf{i}}
\newcommand{\vj}{\, \mathbf{j}}
\newcommand{\vk}{\, \mathbf{k}}
\newcommand{\bi}{\, \mathbf{i}}
\newcommand{\bj}{\, \mathbf{j}}
\newcommand{\bk}{\, \mathbf{k}}
\newcommand{\dist}{\, \mathrm{dist}}
\DeclareMathOperator{\Arg}{\mathrm{Arg}}
\DeclareMathOperator{\Ln}{\mathrm{Ln}}
\newcommand{\imag}{\, \mathrm{i}}

\usepackage{graphicx}
\usepackage{color}
\usepackage{tcolorbox}
\usepackage{tikz}
\usepackage{amsmath}
\usetikzlibrary{arrows}
\usepackage{xcolor}
\shadedsolutions
\definecolor{SolutionColor}{rgb}{0.95,0.95,0.95}

\usepackage{graphicx}
\newcommand\AM{{\sc am}}
\newcommand\PM{{\sc pm}}
     
%\usepackage{twemojis}
\newcommand{\quiz}{15}
\newcommand{\term}{Fall}
\newcommand{\due}{9:55 \AM}
\newcommand{\class}{MATH 202}
\begin{document}



\large
\vspace{0.1in}
\noindent\makebox[3.0truein][l]{\textbf{\class, \term \/ \the\year}}
\textbf{Name:} \hrulefill \\
\noindent \makebox[3.0truein][l]{\textbf{Exam I}}
\textbf{Row and Seat}:\hrulefill\\
%\vspace{0.1in}

\small
\begin{flushleft}
 % \emph{ “Integrity is doing the right thing, even when no one is watching.” \hfill {\sc C.S. Lewis}}
   \emph{“Money buys everything except love, personality, freedom, immortality, silence, peace.”}
    \hfill \sc{Carl Sandburg}
 % “Sometimes it's a little better to travel than to arrive.”}  \hfill {\sc  Robert M. Pirsig}
 %   “Self-education is, I firmly believe, the only kind of education there is.” 
  %  \hfill {\sc  Isaac Asimov }}
\end{flushleft}


\noindent  This exam   has questions 1 through  \numquestions \/ with a total of  \numpoints\/  points.  
\noindent \begin{tcolorbox}
    \begin{minipage}{6.5in}
    \begin{enumerate}
    
    \normalsize 
    \item \textbf{Show all of your work.} Do not expect to earn full credit for a correct answer without the needed work.
    
    \item Divine intervention is \emph{not} a substitute for showing your work.
    
    
    \item If your answer is wrong, but your work shows me that you know the major steps in solving a problem, you will likely earn
    some partial credit.
    
    \item Your work should convince me that not only could you correctly solve the given 
    problem, but you could also solve any related problem.
    
    \item If a question asks for a sentence, write your answer as an English sentence. 
    
    \item No talking, no sharing calculators, and no scratch paper.
    \item  Turn your phone off and put it out of sight.
    \item  Clear your desk of everything, except a pencil, an eraser, and a calculator.
    \item If you never make a mistake, you may use ink; otherwise, use a pencil.
    \item Do not unstaple the pages of your exam.
    \item We'll all start at the same time; it is  the polite thing to do.
    \item Write your answers in the space provided. 
    \item If you do not want something graded, erase it or clearly cross it out. 
    \item You may stare at your feet, your paper, or the ceiling, but nowhere else. 
    \item If you wear a  baseball cap, wear it backwards so I can see your eyes.
    \item Work each problem correctly.
    \item When you are finished, collect your things, place your exam paper in the folder on the front desk,
    and quietly leave the room.
    \item After you turn in your paper, I will not answer questions about the test until after it is graded.
    
    \item Not knowing the rules is not a valid excuse for not following them.
    
    \item Read all directions and problems carefully. 
    \end{enumerate} 
    \end{minipage} 
    
    \end{tcolorbox}
    
\end{document}

    \newpage

    \begin{questions}


\vspace{0.1in}




\question[5] Given that $F$ is a sequence and that $F_1 = 5, F_2 = 7$, and $F_3 = 10$.
Is $F$ an  \emph{arithmetic} sequence? Explain.

%OK
\begin{solution}[2.25in] No, $F$ is not an arithmetic sequence. To be an
  arithmetic sequence, the \emph{difference} of consecutive terms must be constant.
  But we have $F_2 - F_1 \neq F_3 - F_2$, so $F$ is not an arithmetic sequence.

\end{solution}

\question[5] Given that $Q$ is a sequence and that $Q_1 =2, Q_2 = 8$, and $Q_3 = 31$.
Is $Q$ a  \emph{geometric} sequence? Explain.

\begin{solution}%[2.25in] 
No, $Q$ is not a geometric sequence. To be a
  geometric  sequence, the \emph{quotient} of consecutive terms must be constant.
  But we have  $\frac{F_2}{F_1}\neq \frac{F_3}{F_2}$, so $Q$ is not a
  geometric sequence.
\end{solution}

\hfill
\newpage

\question[5] Given that $G$ is an \emph{arithmetic} sequence and that $G_1 = 2$ and $G_2=8$, 
find a \emph{formula} for $G$.

\begin{solution}[2.25in]
  Since $G$  is an \emph{arithmetic} sequence its formula has the form $G_n = a n  + b$,
  where $a$ and $b$ are real numbers. Pasting in the given data, we have
  \begin{align*}
     8 &= 2 a + b, \\
     9 &= 4 a + b.
  \end{align*}
Solving for $a$ and $b$, we have $a=\frac{1}{2}$ and $b=7$. So a formula for $G$ is
$G_n = \frac{1}{2} n + 7$.

\end{solution}

    
\question[5] Given that $H$ is a \emph{geometric} sequence and that $H_1 = 8$ and $H_2=16$, 
find a \emph{formula} for $H$.

\begin{solution}[2.25in]
  Since $H$ is a geometric sequence, its formula has the form $H_n = c a^n$, where $c$ 
  and $a$ are real numbers. Pasting in the given data, we have
  \begin{align*}
      8 &= c a^2 \\
      9 &= c a^4    
  \end{align*}
  Dividing the second equation by the first, we get 
  \begin{equation*}
     \frac{9}{8} = \frac{c a^4}{c a^2} = a^2.
  \end{equation*}
 So $a = \pm \sqrt{\frac{9}{8}}$. We were asked for a formula for $H$, not a representation
 of all possible formulae. So let's be positive and choose $a = \sqrt{\frac{9}{8}}$.
 
 Pasting  in $\sqrt{\frac{9}{8}}$ for $a$ in
 the first equation gives $8= \frac{9}{8} c $. So $c = \frac{64}{9}$ All together,
 \begin{equation*}
     H_n =  \frac{64}{9} \left(\sqrt{\frac{9}{8}}\right)^n.
 \end{equation*}
 Likely we should simplify this to something like
 \begin{equation*}
  H_n =  \frac{64}{9} \left(\frac{9}{8}\right)^{n/2}.
\end{equation*}
But there are forms that are acceptable.
\end{solution}

%OK

\question Find the \emph{numerical value} of each sum
\begin{parts}
  \part[5] $\displaystyle \sum_{k=1}^{12} 1$
  \begin{solution}[1.25in]
  \begin{equation*}
    \sum_{k=1}^{46} 1 = 1 + 1 + \cdots + 1 = 46.
  \end{equation*}
  \end{solution} 

  \part[5] $\displaystyle \sum_{k=1}^{12} (2 k + 1)$
  \begin{solution}%[1.25in]
  \begin{align*}
    \sum_{k=1}^{107} (2 k + 3) &= \left(\sum_{k=1}^{107} 2 k \right)  + 
    \left(\sum_{k=1}^{107} 3 \right) &\mbox{(additivity)} \\
    &= \left(2 \sum_{k=1}^{107} k \right)  + 
    \left(3 \times 107\right) &\mbox{(outative)}\\
    &= 2 \times \frac{107 \times 108}{2} + 3 \times 107 &\mbox{(QRS)}\\
    &= 11877.
  \end{align*}
  \end{solution} 
\end{parts}


\vfill
\newpage
\question[5] After graduation from UNK, you expect a first year salary of $\$42,000$
and you expect a raise of 3\% for each of the 45 years that you plan to work. Your salary
for your $\mbox{n}^{\mbox{th}}$ year of work $S_n$ is
\begin{equation*}
    S_n = 42000 \times 1.03^{n-1}
\end{equation*}
Find your salary in your last year of employment; that is find the numerical value of $S_{45}$.


\begin{solution}[1.25in]
\end{solution}

\question The human population $P$ of Wynot, Nebraska  is an \emph{exponential function} of the years $T$ after the year 2000.  Specifically, the population in the years 2000 and 2010 are given in the table



\begin{figure}[h]
\begin{centering}
\begin{tabular}{|c|c|c|} \hline
  Year & $T$  & $P$ \\ \hline
  2000 & 0    & 250  \\
  2010 & 10   & 230\\ \hline
  \end{tabular}.
  \caption{Human population of Wynot, Nebraska for the years 2000 and 2010.}
  \end{centering}
\end{figure}


\begin{parts}


\part [5] Find the \emph{exponential function} that matches the given data.

\begin{solution}[1.0in] We just need to match to the general result in the QRS. That gives
   \begin{equation*}
       P = 432 \times \left(\frac{425}{432} \right)^{T/10}.
   \end{equation*}
\end{solution}

\part [5] Using your \emph{exponential function} from part `a,' when will the population of Wynot be 210? 

\begin{solution}%[3.0in] 
   We need to solve $250 = 432 \times \left(\frac{425}{432} \right)^{T/10}$ for $T$. Some (many) students will be more comfortable
   using decimal approximations to various quotients as they go. That's OK.
   \begin{align*}
   \left[250 = 432 \times \left(\frac{425}{432} \right)^{T/10} \right] &=
   \left[\frac{250}{432} = \left(\frac{425}{432} \right)^{T/10} \right] &\mbox{($\div$ 432)} \\
   &= \left[\log(\frac{250}{432}) = \log \left(\left(\frac{425}{432} \right)^{T/10} \right) \right] &\mbox{(log of left and right)} \\
   &= \left[\log \left (\frac{250}{432} \right) = \frac{T}{10} \log \left(\frac{425}{432}  \right) \right] &\mbox{(log property)} \\
   &= \left[T = 10 \times \frac{\log \left (\frac{250}{432} \right)}{\log \left(\frac{425}{432} \right)} \right], &\mbox{(divide)} \\
   &= \left[T \approx 334  \mbox{ years} \right]. 
   \end{align*}
This is so far into the future, it's almost surely wildly inaccurate.
\end{solution}

\end{parts}


\hfill
\newpage


\question [5]  Find the solution to the linear equations
\begin{align*}
   2 x + 5 y &= 19, \\
   x -  y &= 1.
\end{align*}

\begin{solution}[3.0in] Let's use our matrix method. Ordering the unknowns as $x$  (first column) and $y$ (second column), 
   our system in matrix form is
   \begin{align*}
       \begin{bmatrix} 5 & 8 & 14 \\ 1 & -1 & 3 \end{bmatrix}
       \begin{bmatrix}  \\  & R_2 \leftarrow R_1 - 5 R_2 \end{bmatrix}
       &= \begin{bmatrix} 5 & 8 & 14 \\ 0 & 13 & -1 \end{bmatrix}
   \end{align*}
   The second equation is $13 y = -1$. So $y=-\frac{1}{13}$. Pasting that into the 
   first equation gives $5 x - \frac{8}{13} = 14$. So $x = \frac{38}{13}$. 

   Should we check our work?  Sure. 
   \begin{equation*}    
      \left[5 x + 8y = 14\right] = \left[5 \times \frac{38}{13} - 8 \times \frac{1}{13} = 14\right] =
     \mbox{True!}      
   \end{equation*}
   And one more time
   \begin{equation*}    
      \left[x - y = 3\right] = \left[\frac{38}{13}  + \frac{1}{13} = 3 \right] 
      = \mbox{True!}      
   \end{equation*}
\end{solution}



\question [5] Define a function $W$ by $W(x) = \sqrt{x}$ and $\dom(W) = [0,16]$. 
    Find the \emph{range} of $W$.
    \begin{solution}[2.5in]
      A graph of the square root function shows that $\range(W) = [0,2].$
      \end{solution}
      
    \question [5] Find the \emph{distance} between the points $(1,5)$ and $(9,8)$.
    \begin{solution}[2.0in]
    We have
    \begin{align*}
        \dist((7,9),(-1,-2)) &= \sqrt{(7+1)^2 + (9 + 2)^2}, &(\mbox{distance formula})\\
                               &= \sqrt{64+ 121}, &(\mbox{arithmetic})\\
                               &= \sqrt{185}. &(\mbox{arithmetic})
    \end{align*}
    \begin{itemize}
    
        \item It's easy to misread a question--be careful. I often put what you need to find in italic type (in this case \emph{distance})  in an attempt to clarify what you are looking for.
        
        \item The factors of 185 are $5$ and $37$.  Neither of these factors are perfect squares, so $\sqrt{185}$
        is properly simplified. 
    
        \item Unless you are asked for a decimal approximation, you should leave
        your answers in an exact form. This problem \emph{doesn't} ask for an exact 
        solution, so  $13.60147$ is \emph{not} a correct solution.
        
        \item A common error is to do the following:
        \begin{align*}
            \sqrt{8^2 + 11^2} = 8 + 11  = 19.
       \end{align*}
        But this is \textbf{wrong}. The equation $\sqrt{x^2 + y^2} = x + y$ is not an identity. You can verify this by 
        pasting in $x = 1$ and $y=1$ into the equation. The result is
         \begin{align*}
            \left[\sqrt{1^2 + 1^2} = 1+ 1 \right]  = \left[\sqrt{2} = 2 \right] = \mbox{False}
       \end{align*}
       Our Quick Reference Sheet lists other common errors.
    \end{itemize}
    
    
    
    \end{solution}
    

    \question[5] The \emph{midpoint} of points $P$ and $(5,7)$ is $(2,3)$. Find the 
    \emph{coordinates} of the point $P$.
    \begin{solution}[2.0in] Let $P = (x,y)$. We have
        \begin{equation*}
            \left(\frac{x+5}{2}, \frac{y+6}{2} \right) = (-2,3).
        \end{equation*}
        So we need to solve
        \begin{align*}
            \frac{x+5}{2} &= -2,\\
            \frac{y+6}{2} &= 3.
        \end{align*}
        for $x$ and for $y$. Solving the first equation for $x$ gives
        \begin{align*}
            \left[  \frac{x+5}{2} = -2 \right] &= [x + 5 = -4], &\mbox{(multiply by 2)}\\
                                                &= [x=-9]. &\mbox{(add -5)}
        \end{align*}
        And solving the second for $y$ gives
        \begin{align*}
            \left[  \frac{y+6}{2} = 3 \right] &= [y+6 = 6], &\mbox{(multiply by 2)}\\
                                               &= [y=0]. &\mbox{(add -6)}
        \end{align*}
        Solving these equations for $x$ and $y$ gives $x=-9$ and $y=0$.
    
    \end{solution}
    \question A line $L$ contains the points $(x=7,y=2)$ and
    $(x=8,y=4)$.  
    
    \begin{parts}
    
        \part [5] Find an \emph{equation} of the line $L$.
        \begin{solution}[1.5in] The slope of the line $L$ is
            \begin{equation*}
                \frac{2-(-1)}{5-7} = -\frac{3}{2}
            \end{equation*}    
            Using the point $(x=5,y=2)$ in the point-slope formula, an equation of
            the line is
            \begin{equation*}
                y-2 = -\frac{3}{2} (x-5).
            \end{equation*}
            Using the other point, gives an equation that looks different, but 
            describes exactly the same  line.
    
            \quad A good way to check our work is to paste
            the data in to the equation and see if it is true. Pasting
            $(x=5,y=2)$ into our equation for $L$, we have
            \begin{equation*}
                \left[2-2 = -\frac{3}{2} (5-5) \right] = 
                   \left[0 = 0\right] = \mbox{True}.
            \end{equation*}
            And pasting in $(x=7,y=-1)$, we have
            \begin{equation*}
                \left[-1-2 = -\frac{3}{2} (7-5) \right] = 
                   \left[-3 = -3 \right] = \mbox{True}.
            \end{equation*}
           
            The question doesn't ask for a particular form for 
            the equation of the line. We've found a point slope form for 
            the line and we have checked our work. What should we do now? We 
            should LIB (let it be).    
            If the question asked for the slope-intercept form, we would need to 
            give the answer as
            \begin{equation*}
                y = -\frac{3}{2}  x + \frac{19}{2}.
            \end{equation*} 
             But the question gives us the \emph{freedom} to give either form for
             the equation of the line.  \textbf{Let's exercise our freedom and do so.}
        \end{solution}
    
        \part  [5] Find the \emph{x-intercept} of the line $L$.
        \begin{solution}[1.5in] To find the x-intercept of the line
            $y-2 = -\frac{3}{2} (x-5)$, set $y$ to zero and solve for $x$; we have
            \begin{align*}
                \left[0-2 = -\frac{3}{2} (x-5) \right] &= 
                    \left[\frac{4}{3}  = x-5 \right],  &\mbox{(multiply by $-2/3$)}\\
                           &= \left[x = \frac{19}{3} \right]. &\mbox{(add 5)}
            \end{align*}
    
    
        \end{solution}
    \end{parts}



\question Find the \emph{vertex} of each parabola

  \begin{parts}


    \part[5] $y - 8 = 5 (x + 2)^2$

    \begin{solution}[1.5in] Matching $y - 8 = \sqrt{2} (x + 2)^2$ to $y - k = a (x-h)^2$
      gives $k=8, h=-2$, and $a=\sqrt{2}$.  
      So the vertex is the point $(-2,8)$.


    \end{solution}

        \part[5] $y=3 {{x}^{2}}-6 x+5$

   
    \begin{solution}%[1.5in] 
    Matching $y=2 {{x}^{2}}-28 x+103$ to
      $y=a x^2 + b x + c$ gives $a=2,b=-28$, and $c=103$. The 
      vertex  is the point
      \begin{equation}
        \left(-\frac{b}{2 a}, c - \frac{b^2}{4 a}\right) = 
        \left(-\frac{-28}{2 \times 2}, 103 - \frac{(-28)^2}{4 \times 2}\right) = 
        \left(7, 5 \right).
      \end{equation}
    \end{solution}

  
  \hfill
  \newpage


  \end{parts}


  \question Follow these steps to solve the inequality $x^2 - x \geq 12$.

  \begin{parts}

    \part [5] \emph{Solve} the equation $x^2 -x = 12$.

    \begin{solution}[2.0in]
      \begin{align*}
        \left[x^2 -x = 12 \right] &= \left[x^2 -x -12 =0 \right]  &\mbox{(subtract 12)}\\
                    &= \left[(x-4)(x+3) =0 \right]  &\mbox{(factor)}\\
                    &= \left[x=4 \mbox{ or } x=-3  \right] & \mbox{(teacher's pet fact)}      
      \end{align*}

    \end{solution}

    \part [5] Check that \emph{both of your solutions are correct} by 
    pasting them into the equation $x^2 -x = 12$.

    \begin{solution}[1.0in]
      Pasting in $x \to 4$ into $x^2 -x = 12$ gives
      \begin{equation*}
          \left[4^2 - 4 = 12\right] = [12=12] = \mbox{True!}
      \end{equation*}
      And Pasting in $x \to -3$ into $x^2 -x = 12$ gives
      \begin{equation*}
          \left[(-3)^2 + 3 = 12\right] = [12=12] = \mbox{True!}
      \end{equation*}

    \end{solution}

    \part [5] Put both of your solutions on a \emph{number line}, correctly
    ordered from \emph{least to greatest}.

    \begin{solution}[1.0in]

      \usetikzlibrary{arrows}
\begin{tikzpicture}
\draw[latex-] (-6.5,0) -- (6.5,0) ;
\draw[-latex] (-6.5,0) -- (6.5,0) ;
\foreach \x in  {-6,-4,-2,0,2,4,6}
\draw[shift={(\x,0)},color=black] (0pt,3pt) -- (0pt,-3pt);
\foreach \x in {-6,-4,-2,0,2,4,6}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-3pt) node[below] 
{$\x$};
%\draw[*-o] (-3,-3) -- (-3,-3)
\draw[*-o] (-3,0);
\draw[*-o] (4,0);
\end{tikzpicture}

    \end{solution}



    \part [5] Make a \emph{table} of the intervals determined by the number line 
    from the previous part, the test points, and the
    value of $x^2 - x \geq 12$ at each test point.
        
    \begin{solution}%[4.0in]

      \begin{tabular}{|c|c|c|c|} \hline \hline 
        \textbf{Interval} &  $\mathbf{x}$  & $\mathbf{x^2 -x \geq 12}$  & \textbf{true or false}\\ \hline 
        $(-\infty,-3)$  & -4 &  $(-4)^2 + 4 \geq 12 $ & true \\  \hline 
        $(-3,4)$  & 0 &  $0^2 + 0 \geq  12 $  &false \\ \hline 
        $(4, \infty)$  & 5 &  $5^2 + 4 \geq 12 $ & true  \\ \hline        
      \end{tabular}

    \end{solution}

    \vfill
    \newpage

    \part [5] Finish the sentence:  The solution set is

    \begin{solution}[1.5in]
    $(-\infty, -3] \cup [4 ,\infty)$. 
    
    We include the endpoints in each interval (use a bracket, not a paren) because we're
    solving $x^2 - x \geq 12$. And that allows equality. Here in 
    a bit, this rule will be modified.

    \end{solution}

  \end{parts}

\question [5] Albert buys a \$1,000 bank certificate of deposit with an APY of 5\%.  How many years must Albert wait until the value of 
the certificate of deposit  is worth \$1,500?

\begin{solution}[2.5in]
\end{solution}

\question  Find the \emph{solution set} of each equation
\begin{parts}

\part [5] $\log(x) = 4$
\begin{solution}[2.5in]
\end{solution}

\part [5] $10^x = -4$
\begin{solution}%[2.5in]
\end{solution}

\hfill
\newpage
\
\part [5] $\log(2 x + 1) = \log(x+4)$
\begin{solution}[2.5in]
\end{solution}

\part[5] $2 = 1.04^x$
\begin{solution}[2.5in]
\end{solution}

\part[5] $2 = (1+x)^8$
\begin{solution}[2.5in]
\end{solution}

\end{parts}

\hfill
\newpage

\question [5] The average surface temperature $T$ of a planet in our solar system  varies \emph{inversely} with the \emph{square root} of
its distance $D$ to the sun.   This means there is a number $k$ such that $T = \frac{k}{\sqrt{D}}$.
Given that the average surface temperature of Mercury is $440^\degree$ Kelvin and that the distance from Mercury to the
sun is 36 million miles, find the average surface temperature of Earth given that the distance from the sun to Earth is 93 million miles. \footnote{Degrees \emph{Kelvin} is just like degrees Celsius,  but  with a different value for zero. The greenhouse effect makes Earth warmer than this simple model indicates.}

\begin{solution}[2.5in]
\end{solution}


\question [5] The formula for a function $Q$ is $Q(x) =  (x-2)^2 + 3$ and its \emph{domain} is $[1,3]$.   Find the \emph{range} of this function.
\begin{solution}[2.5in]
\end{solution}

\end{questions}


\end{document}

  \question Find the solution set to $\frac{2 x + 3}{4 x +1} \leq 1$ by
following these steps.

\begin{parts}

    \part [1] Use algebra tools to find an equivalent inequality of the 
    form $\frac{P(x)}{Q(x)} \leq 0$, where $P$ and $Q$ are polynomials.

    \begin{solution}[1.25in] We have
        \begin{align*}
            \left[\frac{2 x + 3}{4 x +1} \leq 1 \right] &=
               \left[\frac{2 x + 3}{4 x +1} -1 \leq 0\right], & \mbox{(subtract 1)}\\
               & = \left[\frac{2 x + 3}{4 x +1} - \frac{4 x + 1}{4 x +1} \leq 0\right], & \mbox{(make a common denominator)}\\
               &= \left[\frac{-2 x + 2}{4 x +1} \leq 0\right]. & \mbox{(combine numerators)}
        \end{align*}
    
    \end{solution}

    \part[1] Find all x-intercepts and all VAs for $\frac{P(x)}{Q(x)}$.

    \begin{solution}[2.25in] To find the x-intercepts we set the numerator to zero and solve:
        \begin{equation*}
             \left[-2x + 2 = 0\right] = \left[x=1\right].
        \end{equation*}
        To find the VA we set the denominator to zero and solve:
        \begin{equation}
             \left[4x+1 = 0\right] = \left[x= -\frac{1}{4}\right].
        \end{equation}
    
    \end{solution}


    \part [1] Put all x-intercepts and VAs on to a number line.

    \begin{solution}%[1.25in]
        \usetikzlibrary{arrows}
        \begin{tikzpicture}
        \draw[latex-] (-6.5,0) -- (6.5,0) ;
        \draw[-latex] (-6.5,0) -- (6.5,0) ;
        \foreach \x in  {-6,-4,-2,0,2,4,6}
        \draw[shift={(\x,0)},color=black] (0pt,3pt) -- (0pt,-3pt);
        \foreach \x in {-6,-4,-2,0,2,4,6}
        \draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-3pt) node[below] 
        {$\x$};
        \draw[*-o] (-0.25,0);
        \draw[*-o] (1.0,0);
       % \draw[*-o] (0.92,0) -- (2.08,0);
       % \draw[very thick    ] (0.92,0) -- (1.92,0);
        
        \end{tikzpicture}
    \end{solution}

    \vfill 
    \newpage
    \part [1] Build the chart with columns for the interval, the test
    number, evaluation at the test number, and the true/false value.

    \begin{solution}[3.25in]
    
    \begin{tabular}{|c|c|c|c|}
    \hline
     \textbf{Interval} & \textbf{Test} & $\frac{2 x + 3}{4 x +1} \leq 1$ & \textbf{True or False} \\ \hline
     $(-\infty, -1/4)$  & -1  & $-\frac{1}{3} \leq 1$ & True \\
      $(-1/4,1)$  &  0  & $3 \leq 1 $ & False \\
      $(1,\infty)$ & 2 & $\frac{7}{9} \leq 1$ & True \\ \hline
    \end{tabular}
    
    \end{solution}


    \part [1] Test each interval endpoint for inclusion or
    exclusion into the solution set.

    \begin{solution}[3.25in]
    
 \begin{tabular}{|c|c|c|}
 \hline
   \textbf{Endpoint} &  $\frac{2 x + 3}{4 x +1} \leq 1$ & \textbf{True or False} \\ \hline
    $-1/4$   & dne   & False \\
    1           & $1 \leq 1$ & True \\ \hline
 \end{tabular}
    
    \end{solution}

    \part [1] Express the solution set in either interval notation, 
    pictorially, or set builder notation.
    
    \begin{solution}
    In interval notation the solution set is $(-\infty,-\frac{1}{4}) \cup [1,\infty)$. In set builder notation,
    it is $\{x |  x < -\frac{1}{4} \mbox{ or }  x \geq  1 \}$.
    
    \end{solution}

\end{parts}



\question [5]  At 6 \AM,\, Louisa has 340 mg of caffeine circulating 
in her blood. After $T$ hours, the amount of caffeine $C$ in her blood is
\(
     C = 340  \times  0.9^T
\).
When Louisa goes to bed at 10 \PM,\, how much caffeine is
still in circulation?
\begin{solution}[1.5in] We have
    \begin{equation*}
        C = 340 \times 0.9^{16} = \SI{63.0}{\milli\gram}.
    \end{equation*}
I rounded the value to the nearest tenth---given the context, that
is reasonable.
\end{solution}


\question Find the inverse of the function $f(x) = \frac{2 x + 1}{x-1}, \,\, x \neq 1$.
\begin{solution}[2.5in]
    We need to solve $y = \frac{2 x + 1}{x-1}$ for $x$. 
    When $y = 2$, the solution set is empty; otherwise, the
    solution is 
    \begin{equation*}
         x = \frac{y+1}{y-2}                           
    \end{equation*}
    So $f^{-1}(y) = \frac{y+1}{y-2}$ and $\dom(f^{-1}) = \{y | y \neq 2 \}$.
        
 
\end{solution}\question Define a function $W$ by $W(x) = \sqrt{x}$ and $\dom(W) = [0,4]$. 
    Find the \emph{range} of $W$.
    \begin{solution}
      A graph of the square root function shows that $\range(W) = [0,2].$
      \end{solution}
    \question [5] Find the \emph{distance} between the points $(7,9)$ and $(-1,-2)$.
    \begin{solution}[2.0in]
    We have
    \begin{align*}
        \dist((7,9),(-1,-2)) &= \sqrt{(7+1)^2 + (9 + 2)^2}, &(\mbox{distance formula})\\
                               &= \sqrt{64+ 121}, &(\mbox{arithmetic})\\
                               &= \sqrt{185}. &(\mbox{arithmetic})
    \end{align*}
    \begin{itemize}
    
        \item It's easy to misread a question--be careful. I often put what you need to find in italic type (in this case \emph{distance})  in an attempt to clarify what you are looking for.
        
        \item The factors of 185 are $5$ and $37$.  Neither of these factors are perfect squares, so $\sqrt{185}$
        is properly simplified. 
    
        \item Unless you are asked for a decimal approximation, you should leave
        your answers in an exact form. This problem \emph{doesn't} ask for an exact 
        solution, so  $13.60147$ is \emph{not} a correct solution.
        
        \item A common error is to do the following:
        \begin{align*}
            \sqrt{8^2 + 11^2} = 8 + 11  = 19.
       \end{align*}
        But this is \textbf{wrong}. The equation $\sqrt{x^2 + y^2} = x + y$ is not an identity. You can verify this by 
        pasting in $x = 1$ and $y=1$ into the equation. The result is
         \begin{align*}
            \left[\sqrt{1^2 + 1^2} = 1+ 1 \right]  = \left[\sqrt{2} = 2 \right] = \mbox{False}
       \end{align*}
       Our Quick Reference Sheet lists other common errors.
    \end{itemize}
    
    
    
    \end{solution}
    
    \question[5] The \emph{midpoint} of points $P$ and $(5,6)$ is $(-2,3)$. Find the 
    \emph{coordinates} of the point $P$.
    \begin{solution}[2.0in] Let $P = (x,y)$. We have
        \begin{equation*}
            \left(\frac{x+5}{2}, \frac{y+6}{2} \right) = (-2,3).
        \end{equation*}
        So we need to solve
        \begin{align*}
            \frac{x+5}{2} &= -2,\\
            \frac{y+6}{2} &= 3.
        \end{align*}
        for $x$ and for $y$. Solving the first equation for $x$ gives
        \begin{align*}
            \left[  \frac{x+5}{2} = -2 \right] &= [x + 5 = -4], &\mbox{(multiply by 2)}\\
                                                &= [x=-9]. &\mbox{(add -5)}
        \end{align*}
        And solving the second for $y$ gives
        \begin{align*}
            \left[  \frac{y+6}{2} = 3 \right] &= [y+6 = 6], &\mbox{(multiply by 2)}\\
                                               &= [y=0]. &\mbox{(add -6)}
        \end{align*}
        Solving these equations for $x$ and $y$ gives $x=-9$ and $y=0$.
    
    \end{solution}
    \question A line $L$ contains the points $(x=5,y=2)$ and
    $(x=7,y=-1)$.  
    
    \begin{parts}
    
        \part [1] Find an \emph{equation} of the line $L$.
        \begin{solution}[1.5in] The slope of the line $L$ is
            \begin{equation*}
                \frac{2-(-1)}{5-7} = -\frac{3}{2}
            \end{equation*}    
            Using the point $(x=5,y=2)$ in the point-slope formula, an equation of
            the line is
            \begin{equation*}
                y-2 = -\frac{3}{2} (x-5).
            \end{equation*}
            Using the other point, gives an equation that looks different, but 
            describes exactly the same  line.
    
            \quad A good way to check our work is to paste
            the data in to the equation and see if it is true. Pasting
            $(x=5,y=2)$ into our equation for $L$, we have
            \begin{equation*}
                \left[2-2 = -\frac{3}{2} (5-5) \right] = 
                   \left[0 = 0\right] = \mbox{True}.
            \end{equation*}
            And pasting in $(x=7,y=-1)$, we have
            \begin{equation*}
                \left[-1-2 = -\frac{3}{2} (7-5) \right] = 
                   \left[-3 = -3 \right] = \mbox{True}.
            \end{equation*}
           
            The question doesn't ask for a particular form for 
            the equation of the line. We've found a point slope form for 
            the line and we have checked our work. What should we do now? We 
            should LIB (let it be).    
            If the question asked for the slope-intercept form, we would need to 
            give the answer as
            \begin{equation*}
                y = -\frac{3}{2}  x + \frac{19}{2}.
            \end{equation*} 
             But the question gives us the \emph{freedom} to give either form for
             the equation of the line.  \textbf{Let's exercise our freedom and do so.}
        \end{solution}
    
        \part  [1] Find the \emph{x-intercept} of the line $L$.
        \begin{solution}[1.5in] To find the x-intercept of the line
            $y-2 = -\frac{3}{2} (x-5)$, set $y$ to zero and solve for $x$; we have
            \begin{align*}
                \left[0-2 = -\frac{3}{2} (x-5) \right] &= 
                    \left[\frac{4}{3}  = x-5 \right],  &\mbox{(multiply by $-2/3$)}\\
                           &= \left[x = \frac{19}{3} \right]. &\mbox{(add 5)}
            \end{align*}
    
    
        \end{solution}
    \end{parts}


    \question [5] The number of lawns $L$ a work crew can mow in a day 
varies jointly with the number of people $N$ in the crew and with
the time $T$  they work in a day.

Given that $L = 8$ when $N = 5$ and $T = 6$, find $L$ when
$N = 8$ and $T = 10$.
\begin{solution}[3.5in]
There is a constant $k$ such that $L =k N T$. This formula should make sense--doubling the size of the work crew, we
should be able to mow twice as many lawns--the formula shows that is true. The same is true for working twice as long.

Pasting in $L = 8$ when $N = 5$ and $T = 6$ into $L =k N T$ yields $8 = 30 k$.  So $k = \frac{4}{15}$.  That makes
our formula
\begin{equation*}
L = \frac{4}{15} N T.
\end{equation*}
Pasting in $N = 8$ and $T = 10$ gives
\begin{equation*}
L = \frac{4}{15} \times 80 = \frac{64}{3} \approx 21.3.
\end{equation*}


\textbf{Teacher's embellishment} A century or so ago, students were taught 
a short-cut method for such problems called the \emph{double rule of three}.
The Lewis Carroll poem ``The Mad Gardener's Song'' refers to this method.


\end{solution}

\question Find the \emph{vertex} of each parabola

  \begin{parts}

    \part[5] $y - 8 = x^2$

    \begin{solution}[1.5in] Matching $y - 8 = x^2$ to $y - k = a (x-h)^2$
      gives $k=8, h=0$, and $a=1$.  So the vertex is the point $(0,8)$.

    \end{solution}

    \part[5] $y - 8 = \sqrt{2} (x + 2)^2$

    \begin{solution}[1.5in] Matching $y - 8 = \sqrt{2} (x + 2)^2$ to $y - k = a (x-h)^2$
      gives $k=8, h=-2$, and $a=\sqrt{2}$.  
      So the vertex is the point $(-2,8)$.


    \end{solution}

        \part[5] $y=2 {{x}^{2}}-28 x+103$

   
    \begin{solution}[1.5in] Matching $y=2 {{x}^{2}}-28 x+103$ to
      $y=a x^2 + b x + c$ gives $a=2,b=-28$, and $c=103$. The 
      vertex  is the point
      \begin{equation}
        \left(-\frac{b}{2 a}, c - \frac{b^2}{4 a}\right) = 
        \left(-\frac{-28}{2 \times 2}, 103 - \frac{(-28)^2}{4 \times 2}\right) = 
        \left(7, 5 \right).
      \end{equation}
    \end{solution}

    \part[5] $y=3 (x-2)(x-4)$
    \begin{solution} To match $y=3 (x-2)(x-4)$ to $y=a x^2 + b x + c$,
      we need to expand (use FOIL) $y=3 (x-2)(x-4)$. We have
      \begin{equation*}
          y = 3 (x-2)(x-4) = 3(x^2 -6 x + 8) = 3 x^2 - 18 x + 24.
      \end{equation*}
      So $a=3, b=-18$, and $c=24$. The 
      vertex  is the point
      \begin{equation*}
        \left(-\frac{b}{2 a}, c - \frac{b^2}{4 a}\right) = 
        \left(-\frac{-18}{2 \times 3}, 24 -\frac{(-18)^2}{4 \times 3} \right) =
        \left( 3, -3 \right).
      \end{equation*}
    \end{solution}


  \end{parts}


  \question Follow these steps to solve the inequality $x^2 - x \geq 12$.

  \begin{parts}

    \part [5] \emph{Solve} the equation $x^2 -x = 12$.

    \begin{solution}[2.0in]
      \begin{align*}
        \left[x^2 -x = 12 \right] &= \left[x^2 -x -12 =0 \right]  &\mbox{(subtract 12)}\\
                    &= \left[(x-4)(x+3) =0 \right]  &\mbox{(factor)}\\
                    &= \left[x=4 \mbox{ or } x=-3  \right] & \mbox{(teacher's pet fact)}      
      \end{align*}

    \end{solution}

    \part [2] Check that \emph{both of your solutions are correct} by 
    pasting them into the equation $x^2 -x = 12$.

    \begin{solution}[2.0in]
      Pasting in $x \to 4$ into $x^2 -x = 12$ gives
      \begin{equation*}
          \left[4^2 - 4 = 12\right] = [12=12] = \mbox{True!}
      \end{equation*}
      And Pasting in $x \to -3$ into $x^2 -x = 12$ gives
      \begin{equation*}
          \left[(-3)^2 + 3 = 12\right] = [12=12] = \mbox{True!}
      \end{equation*}

    \end{solution}

    \part [2] Put both of your solutions on a \emph{number line}, correctly
    ordered from \emph{least to greatest}.

    \begin{solution}%[1.0in]

      \usetikzlibrary{arrows}
\begin{tikzpicture}
\draw[latex-] (-6.5,0) -- (6.5,0) ;
\draw[-latex] (-6.5,0) -- (6.5,0) ;
\foreach \x in  {-6,-4,-2,0,2,4,6}
\draw[shift={(\x,0)},color=black] (0pt,3pt) -- (0pt,-3pt);
\foreach \x in {-6,-4,-2,0,2,4,6}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-3pt) node[below] 
{$\x$};
%\draw[*-o] (-3,-3) -- (-3,-3)
\draw[*-o] (-3,0);
\draw[*-o] (4,0);
\end{tikzpicture}

    \end{solution}

  %  \vfill
  %  \newpage

    \part [2] Make a \emph{table} of the intervals determined by the number line 
    from the previous part, the test points, and the
    value of $x^2 - x \geq 12$ at each test point.
        
    \begin{solution}[4.0in]

      \begin{tabular}{|c|c|c|c|} \hline \hline 
        \textbf{Interval} &  $\mathbf{x}$  & $\mathbf{x^2 -x \geq 12}$  & \textbf{true or false}\\ \hline 
        $(-\infty,-3)$  & -4 &  $(-4)^2 + 4 \geq 12 $ & true \\  \hline 
        $(-3,4)$  & 0 &  $0^2 + 0 \geq  12 $  &false \\ \hline 
        $(4, \infty)$  & 5 &  $5^2 + 4 \geq 12 $ & true  \\ \hline        
      \end{tabular}

    \end{solution}


    \part [2] Finish the sentence:  The solution set is

    \begin{solution}
    $(-\infty, -3] \cup [4 ,\infty)$. 
    
    We include the endpoints in each interval (use a bracket, not a paren) because we're
    solving $x^2 - x \geq 12$. And that allows equality. Here in 
    a bit, this rule will be modified.

    \end{solution}

  \end{parts}

  \question Find the solution set to $\frac{2 x + 3}{4 x +1} \leq 1$ by
following these steps.

\begin{parts}

    \part [1] Use algebra tools to find an equivalent inequality of the 
    form $\frac{P(x)}{Q(x)} \leq 0$, where $P$ and $Q$ are polynomials.

    \begin{solution}[1.25in] We have
        \begin{align*}
            \left[\frac{2 x + 3}{4 x +1} \leq 1 \right] &=
               \left[\frac{2 x + 3}{4 x +1} -1 \leq 0\right], & \mbox{(subtract 1)}\\
               & = \left[\frac{2 x + 3}{4 x +1} - \frac{4 x + 1}{4 x +1} \leq 0\right], & \mbox{(make a common denominator)}\\
               &= \left[\frac{-2 x + 2}{4 x +1} \leq 0\right]. & \mbox{(combine numerators)}
        \end{align*}
    
    \end{solution}

    \part[1] Find all x-intercepts and all VAs for $\frac{P(x)}{Q(x)}$.

    \begin{solution}[2.25in] To find the x-intercepts we set the numerator to zero and solve:
        \begin{equation*}
             \left[-2x + 2 = 0\right] = \left[x=1\right].
        \end{equation*}
        To find the VA we set the denominator to zero and solve:
        \begin{equation}
             \left[4x+1 = 0\right] = \left[x= -\frac{1}{4}\right].
        \end{equation}
    
    \end{solution}


    \part [1] Put all x-intercepts and VAs on to a number line.

    \begin{solution}%[1.25in]
        \usetikzlibrary{arrows}
        \begin{tikzpicture}
        \draw[latex-] (-6.5,0) -- (6.5,0) ;
        \draw[-latex] (-6.5,0) -- (6.5,0) ;
        \foreach \x in  {-6,-4,-2,0,2,4,6}
        \draw[shift={(\x,0)},color=black] (0pt,3pt) -- (0pt,-3pt);
        \foreach \x in {-6,-4,-2,0,2,4,6}
        \draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-3pt) node[below] 
        {$\x$};
        \draw[*-o] (-0.25,0);
        \draw[*-o] (1.0,0);
       % \draw[*-o] (0.92,0) -- (2.08,0);
       % \draw[very thick    ] (0.92,0) -- (1.92,0);
        
        \end{tikzpicture}
    \end{solution}

    \vfill 
    \newpage
    \part [1] Build the chart with columns for the interval, the test
    number, evaluation at the test number, and the true/false value.

    \begin{solution}[3.25in]
    
    \begin{tabular}{|c|c|c|c|}
    \hline
     \textbf{Interval} & \textbf{Test} & $\frac{2 x + 3}{4 x +1} \leq 1$ & \textbf{True or False} \\ \hline
     $(-\infty, -1/4)$  & -1  & $-\frac{1}{3} \leq 1$ & True \\
      $(-1/4,1)$  &  0  & $3 \leq 1 $ & False \\
      $(1,\infty)$ & 2 & $\frac{7}{9} \leq 1$ & True \\ \hline
    \end{tabular}
    
    \end{solution}


    \part [1] Test each interval endpoint for inclusion or
    exclusion into the solution set.

    \begin{solution}[3.25in]
    
 \begin{tabular}{|c|c|c|}
 \hline
   \textbf{Endpoint} &  $\frac{2 x + 3}{4 x +1} \leq 1$ & \textbf{True or False} \\ \hline
    $-1/4$   & dne   & False \\
    1           & $1 \leq 1$ & True \\ \hline
 \end{tabular}
    
    \end{solution}

    \part [1] Express the solution set in either interval notation, 
    pictorially, or set builder notation.
    
    \begin{solution}
    In interval notation the solution set is $(-\infty,-\frac{1}{4}) \cup [1,\infty)$. In set builder notation,
    it is $\{x |  x < -\frac{1}{4} \mbox{ or }  x \geq  1 \}$.
    
    \end{solution}

\end{parts}

\question [2]  At 6 \AM,\, Louisa has 340 mg of caffeine circulating 
in her blood. After $T$ hours, the amount of caffeine $C$ in her blood is
\(
     C = 340  \times  0.9^T
\).
When Louisa goes to bed at 10 \PM,\, how much caffeine is
still in circulation?
\begin{solution}[1.5in] We have
    \begin{equation*}
        C = 340 \times 0.9^{16} = \SI{63.0}{\milli\gram}.
    \end{equation*}
I rounded the value to the nearest tenth---given the context, that
is reasonable.
\end{solution}


\question Find the inverse of the function $f(x) = \frac{2 x + 1}{x-1}, \,\, x \neq 1$.
\begin{solution}[2.5in]
    We need to solve $y = \frac{2 x + 1}{x-1}$ for $x$. 
    When $y = 2$, the solution set is empty; otherwise, the
    solution is 
    \begin{equation*}
         x = \frac{y+1}{y-2}                           
    \end{equation*}
    So $f^{-1}(y) = \frac{y+1}{y-2}$ and $\dom(f^{-1}) = \{y | y \neq 2 \}$.
        
 
\end{solution}


%\includepdf[pages=-]{college-algebra-quick-reference}


   \question [5] The number of lawns $L$ a work crew can mow in a day 
varies jointly with the number of people $N$ in the crew and with
the time $T$  they work in a day.

Given that $L = 8$ when $N = 5$ and $T = 6$, find $L$ when
$N = 8$ and $T = 10$.
\begin{solution}[3.5in]
There is a constant $k$ such that $L =k N T$. This formula should make sense--doubling the size of the work crew, we
should be able to mow twice as many lawns--the formula shows that is true. The same is true for working twice as long.

Pasting in $L = 8$ when $N = 5$ and $T = 6$ into $L =k N T$ yields $8 = 30 k$.  So $k = \frac{4}{15}$.  That makes
our formula
\begin{equation*}
L = \frac{4}{15} N T.
\end{equation*}
Pasting in $N = 8$ and $T = 10$ gives
\begin{equation*}
L = \frac{4}{15} \times 80 = \frac{64}{3} \approx 21.3.
\end{equation*}


\textbf{Teacher's embellishment} A century or so ago, students were taught 
a short-cut method for such problems called the \emph{double rule of three}.
The Lewis Carroll poem ``The Mad Gardener's Song'' refers to this method.


\end{solution}
